#Lab_03 {lab_03}

rob wells, 8/14/2021

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here

# Load janitor here


```

### Load data

In this data set, each row represents a city or town in Maryland with the 2010 and 2020 figures for:

-   Total population
-   White only population
-   Black or African American only population
-   Hispanic only population
-   Native American only population
-   Asian only population
-   Pacific Islander (including native Hawiian) only population

Background on this data from [Social Explorer](https://www.socialexplorer.com/data/ACS2016_5yr/metadata/?ds=ACS16_5yr&table=B02001)

```{r}
# Write code to load the Maryland race by city, place for 2010 - 2020 and call it md_race.
# The data is called lab_03.csv, and is in the same folder as lab_03.qmd (this data notebook)

# md_race <- rio::import("lab_03.csv", which = "md_race")

```

## Answer questions

Q1: Which five cities have the largest overall Hispanic population in 2020? Name the five cities that had the largest Hispanic population in 2020.

A1:

```{r}

# 
md_race %>% 
  select(place, x2020_hispanic) %>% 
  slice_max(x2020_hispanic, n = 5)
```  

# Baltimore City, Wheaton CDP, Germantown CDP, Silver Spring CDP, Aspen Hill CDP

```

Q2: Find the cities with the largest Black population and arrange the results in descending order.

A2:

```{r}
```{r}
md_race %>% 
  select(place, x2020_black) %>%
  group_by(place) %>%
  arrange(desc(x2020_black))
```
# Top 10: Baltimore City, Waldorf CDP, Bowie City, Clinton CDP, Columbia CDP, Randallstown CDP, Milford Mill CDP, Silver Spring CDP, Woodlawn CDP, Suitland CDP



Q3: Create a table called top_hispanic that shows the top 15 communities with the largest Hispanic population in 2020, and also display the 2010 Hispanic population and the total community population for 2020. Look at your results - what stands out to you? Craft a one-paragraph story pitch for your editor.

A3:


```{r}
top_hispanic <- md_race %>% 
  select(place, x2020_hispanic, x2010_hispanic) %>% 
  slice_max(x2020_hispanic, n = 15)
  
```
# All of the listed neighborhoods showed a substantial increase in their Hispanic populations, with Baltimore City's hispanic population increasing by around 10,000 people from 2010 to 2020. The lowest increase on the table was still in the thousands. A story could be that every single neighborhood with the 15 highest Hispanic populations experienced an increase, and Baltimore City's was especially high.


