---
title: "Lawson_assignment_03"
output: html_document
date: "2024-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **An Analysis of Crime Rates in Lynchburg, Virginia from 2012 to 2023**

## Harper Lawson

### Background

XXX

The online data can be found [here](https://www.lynchburgvapolice.gov/crime/).

### Process

To load the necessary packages:

```{r}

library (tidyverse)
library (tidycensus)
library (rmarkdown)
library (janitor)
library (ggplot2)
library (ggthemes)
library (dplyr)

# Assuming these packages are already installed.
# Assuming base, datasets, forcats, graphics, purrr, readr, and stats are already installed and loaded.

```

To load the necessary data:

```{r}

crime_type_lb <- read_csv("data/crime_type_lynchburg.csv") %>%
  clean_names()
general_crime_lb <- read_csv("data/general_crime_lynchburg.csv") %>%
  clean_names()
arrests_lb <- read_csv("data/arrests_lynchburg.csv") %>%
  clean_names()
police_assaults_lb <- read_csv("data/police_assault_lynchburg.csv") %>%
  clean_names()

```

### Question 1:

#### In which areas has Lynchburg seen a *decrease* in crime? In which areas has Lynchburg seen an *increase* in crime?

### Question 2:

#### Out of all crime in Lynchburg, what trends have there been in the violent crime rate?

### Question 3:

#### Which year saw the most crime? The least?

### Question 4:

#### What is the most common crime committed in Lynchburg?

### Question 5:

#### How do each of these findings compare to the state of Virginia as a whole?
