---
title: "Lawson_assignment_03"
output: html_document
date: "2024-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **An Analysis of Crime Rates in Lynchburg, Virginia from 2012 to 2023**

## Harper Lawson

### Background

This data notebook includes data from the Lynchburg Police Department, the Federal Bureau of Investigation and the U.S. Census Bureau. The data ranges from 2012-2023.

Analysis includes that of Lynchburg's population, violent crime rate, arrests by crime, general crime report, and assaults on police officers. Also analyzed is the state of Virginia's violent crime rate.

The online data from the Lynchburg Police Department can be found [here](https://www.lynchburgvapolice.gov/crime/).

The online data from the FBI can be found [here](https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/explorer/crime/crime-trend).

The online data from the U.S. Census can be found [here](https://www.census.gov/quickfacts/fact/table/lynchburgcitycountyvirginia/PST045223).

### Process

To load the necessary packages:

```{r}

library (tidyverse)
library (tidycensus)
library (rmarkdown)
library (janitor)
library (ggplot2)
library (ggthemes)
library (dplyr)

# Assuming these packages are already installed.
# Assuming base, datasets, forcats, graphics, purrr, readr, and stats are already installed and loaded.

```

To load the necessary data:

```{r}

crime_type_lb <- read_csv("data/crime_type_lynchburg.csv") %>%
  clean_names()
general_crime_lb <- read_csv("data/general_crime_lynchburg.csv") %>%
  clean_names()
arrests_lb <- read_csv("data/arrests_lynchburg.csv") %>%
  clean_names()
police_assaults_lb <- read_csv("data/police_assault_lynchburg.csv") %>%
  clean_names()
lb_pop <- read_csv("data/lynchburg_pop.csv") %>%
  clean_names()
va_crime <- read_csv("data/crime_rates_va.csv") %>%
  clean_names()

```

### Question 1:

#### In which areas has Lynchburg seen a *decrease* in crime? In which areas has Lynchburg seen an *increase* in crime?

### Question 2:

#### Out of all crime in Lynchburg, what trends have there been in the violent crime rate?

### Question 3:

#### How does Lynchburg's violent crime rate compare to that of the state of Virginia?

### Question 4:

#### Which year saw the most crime? The least?

### Question 5:

#### What is the most common crime committed in Lynchburg?
