---
editor: 
  markdown: 
    wrap: sentence
---

#lab_08 Sean Mussenden, Aug 16, 2022

## About this lab

To complete this lab, you need to: \* write code in empty codeblocks provided to answer questions included (look for **Q**).
\* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task**: Load janitor and the tidyverse

## Load data

**Task**: Load four tables of U.S.
Census data.
Each table has one row per Maryland county with 2020 population information.
Each table has these same two columns:

-   geoid - a unique identifier for each county
-   name - the county name

The third column in each table represents the total population in one of four demographic categories.

-   maryland_population.csv - total_pop - total population
-   maryland_hispanic.csv - hispanic_pop - total Hispanic population
-   maryland_white.csv - white_pop - total White alone population
-   maryland_black.csv - black_pop - total Black alone population

```{r}

total_pop_df <- read_csv("data/maryland_population.csv")
hispanic_pop_df <- read_csv("data/maryland_hispanic.csv")
white_pop_df <- read_csv("data/maryland_white.csv")
black_pop_df <- read_csv("data/maryland_black.csv")

```

Some of the information in the tables may need to be cleaned or validated in order to produce the correct answer.

## Questions

**Q1** What Maryland county has the largest Hispanic population (by percentage of total population)?

```{r}

hispanic_pop_df <- hispanic_pop_df %>%
  mutate(name = case_when(
    name == "Montgomery County,Maryland" ~ "Montgomery County, Maryland",
    TRUE ~ name
  ))

hispanic_pop_df <- hispanic_pop_df %>%
  inner_join(total_pop_df, by = c("geoid", "name")) %>%
  mutate(percentage = (hispanic_pop / total_pop) * 100)

hispanic_pop_df %>%
  arrange(desc(percentage)) %>%
  head(n=1)
  
  
```

**A1**

Montgomery County, MD has the highest Hispanic population by percentage of total population.

**Q2** What is the total population of people who identify as white alone in Maryland?

```{r}

white_pop_df <- white_pop_df %>%
  mutate(white_pop = as.character(white_pop)) %>%
  mutate(white_pop = case_when(
    white_pop == "13510000" ~ "13510",
    TRUE ~ white_pop
  ))
  sum(as.numeric(white_pop_df$white_pop))


```

**A2**

3,275,048 people identify as white alone in Maryland.

**Q3** What is the average Black population in all Maryland counties?

```{r}


black_pop_df <- black_pop_df %>%
  mutate(black_pop = case_when(
    name == "Baltimore County, Maryland" ~ 243098,
    name == "Baltimore city, Maryland" ~ 375002,
    TRUE ~ black_pop
  ))

summary(black_pop_df$black_pop)

```

**A3**

The average Black population in all Maryland counties is 75,130.
