---
title: "Assignment 1"
output: html_notebook
---

To load the necessary packages:

```{r}

library (tidyverse)
library (rmarkdown)
library (janitor)
library (ggplot2)
library (ggthemes)
library (dplyr)

# Assuming these packages are already installed.
# Assuming base, datasets, forcats, graphics, purrr, readr, and stats are already installed and loaded.

```

To load and view the necessary data:

```{r}

balt_pop <- read_csv("balt_pop_16_19.csv")
view(balt_pop)

```

For each year, the values below are the percent of 16-19-year-olds in Baltimore who are either enrolled in school or employed.

Question 1:
Which 5 communities in Baltimore experienced the highest growth of 16-19 year olds who were either employed or enrolled in school? Which 5 experienced the least growth (largest decrease)? What was the percent change of each?

```{r}

growth <- balt_pop %>% #Creates new data frame
  select(Place_1, Place_2, x2010, x2020) %>% #Selects which columns will be in the new data frame
  mutate(pct_change = ((x2020-x2010)/x2010)*100) #Adds column for percent change and calculates percent change

```

To calculate the 5 communities with the highest growth:
```{r}
growth %>%
  group_by(Place_1) %>%
  summarise(
    pct_change
  ) %>% 
  arrange(desc(pct_change)) %>%
  head(5)
```
To calculate the 5 communities with the least growth:
```{r}
growth %>%
  group_by(Place_1) %>%
  summarise(
    pct_change
  ) %>% 
  arrange(pct_change) %>%
  head(5)
```



