---
title: "lawson_assignment_01"
output: html_notebook
---

# **An Analysis of Trends in Baltimore's 16-19-year-old teens**

## Harper Lawson

### Background

The data utilized in this notebook is from the Vital Signs Open Data Portal. The source is the American Community Survey, a yearly demographics survey administered by the U.S. Census Bureau. The title of the spreadsheet online is "Percentage of Population aged 16-19 in School and/or Employed."

It collects data from 2010-2020. It is worth noting that the boundaries of certain neighborhoods underwent small changes before the data was collected in 2020, which is why there is a separate column with updated neighborhood names for 2020.

It can be inferred that if teenagers are neither enrolled in school or working a job, that their financial situation is not ideal. This is why this data is a fair representation of wealth inequality in Baltimore, as the neighborhoods have varying statistics.

The online data can be found [here](https://vital-signs-bniajfi.hub.arcgis.com/maps/7d814d51552b45e2bfe17e4edf417131/about){.uri}.

### Process

To load the necessary packages:

```{r}

library (tidyverse)
library (rmarkdown)
library (janitor)
library (ggplot2)
library (ggthemes)
library (dplyr)

# Assuming these packages are already installed.
# Assuming base, datasets, forcats, graphics, purrr, readr, and stats are already installed and loaded.

```

To load the necessary data:

```{r}

balt_pop <- read_csv("balt_pop_16_19.csv")

```

For each year, the values below are the percent of 16-19-year-olds in Baltimore who are either enrolled in school or employed.

#### Question 1:

**A)** Which 5 communities in Baltimore experienced the highest growth of 16-19 year olds who were either employed or enrolled in school? What was the percent change of each?

To make a new data frame with growth info:

```{r}

growth <- balt_pop %>%
  select(Place_1, Place_2, x2010, x2020) %>%
  mutate(pct_change = round(((x2020-x2010)/x2010)*100, 2))

```

To calculate the 5 communities with the highest growth:

```{r}

growth %>% 
  select(Place_1, pct_change) %>% 
  slice_max(pct_change, n=5)

```

**B)** South Baltimore had the highest growth by a landslide. What percentage of 16-19-year-olds were in school or employed in 2010? 2015? 2020?

```{r}

south_balt <- subset(balt_pop, Place_1 == "South Baltimore")

south_balt %>%  
select(Place_1, x2010, x2015, x2020)

```

By looking at my new south_balt table, I can see that 47.92% of 16-19-year-olds were in school or employed in 2010, 98.48% in 2015, and 100.00% in 2020. Let's see what the overall growth looks like.

```{r}

south_balt %>%
  select(x2010, x2011, x2012, x2013, x2014, x2015, x2016, x2017, x2018, x2019, x2020) %>%
  gather(key = "year", value = "population", x2010:x2020) %>%
  ggplot() +
  geom_bar(aes(x = year, y = population, fill = population), stat = "identity") +
  labs(x = "Year", y = "Population", title = "Number of 16-19-year-olds in school or employed in South Baltimore (2010-2020")


```

Looking at the chart, was this growth gradual, or did it happen quickly over a short time period?

#### Question 2:

Which one-year period showed the highest growth in South Baltimore?

```{r}

south_balt <- south_balt %>%
  mutate(pct_change_2010_2011 = round(((x2011-x2010)/x2010)*100, 2)) %>%
  mutate(pct_change_2011_2012 = round(((x2012-x2011)/x2011)*100, 2)) %>%
  mutate(pct_change_2012_2013 = round(((x2013-x2012)/x2012)*100, 2)) %>%
  mutate(pct_change_2013_2014 = round(((x2014-x2013)/x2013)*100, 2)) %>%
  mutate(pct_change_2014_2015 = round(((x2015-x2014)/x2014)*100, 2)) %>%
  mutate(pct_change_2015_2016 = round(((x2016-x2015)/x2015)*100, 2)) %>%
  mutate(pct_change_2016_2017 = round(((x2017-x2016)/x2016)*100, 2)) %>%
  mutate(pct_change_2017_2018 = round(((x2018-x2017)/x2017)*100, 2)) %>%
  mutate(pct_change_2018_2019 = round(((x2019-x2018)/x2018)*100, 2)) %>%
  mutate(pct_change_2019_2020 = round(((x2020-x2019)/x2019)*100, 2))

south_balt %>%  
  select(pct_change_2010_2011, pct_change_2011_2012, pct_change_2012_2013, pct_change_2013_2014, pct_change_2014_2015, pct_change_2015_2016, pct_change_2016_2017, pct_change_2017_2018, pct_change_2018_2019, pct_change_2019_2020)

```

As we can see from this table, 2012-2013 experienced the highest growth of 16-19-year-olds in school or employed at 83.83%.

#### Question 3:

**A)** Looking at the south_balt table, when was the percentage of 16-19-year-olds in school or employed in South Baltimore at or below 50%?

```{r}

south_balt1 <- south_balt %>% 
  select(Place_1, x2010, x2011, x2012, x2013, x2014, x2015, x2016, x2017, x2018, x2019, x2020) %>% 
  pivot_longer(names_to="year",
               values_to="percent",
               cols =2:10) %>% 
  select(Place_1, year, percent) 

south_balt1 %>% 
  filter(percent <= 50)

```

We can see from this table that the percentage of 16-19-year-olds in school or employed in South Baltimore was at or below 50% in 2010, 2011, and 2012.

**B)** How do these values compare to Baltimore's averages during these years?

First, I will organize the mean values of 2010, 2011, and 2012.

```{r}

  balt_means <- colMeans(select(balt_pop, x2010, x2011, x2012), na.rm = TRUE)
  tibble(Year = c("x2010", "x2011", "x2012"), Mean = balt_means)

```

Now, I will look at the South Baltimore values again on their own...

```{r}

south_balt %>%
  select(x2010, x2011, x2012)

```

...assign their values and the mean values...

```{r}

#2010
south_balt_2010 <- 47.92
mean_2010 <- 84.73

#2011
south_balt_2011 <- 50.00
mean_2011 <- 85.30

#2012
south_balt_2012 <- 37.04
mean_2012 <- 86.00

```

...and do some subtraction (mean value minus South Baltimore value of the same year).

```{r}
mean_2010-south_balt_2010
mean_2011-south_balt_2011
mean_2012-south_balt_2012
```

And now we see how far South Baltimore's 2010-12 percentages were below Baltimore's mean values.

### Conclusion

2010-2012 was a major low point for the community of South Baltimore compared to Baltimore as a whole in terms of how many 16-19-year-olds were in school or employed. On the other hand, 2012-2013 showed amazing growth.
